---
title: "Crime Rates"
---

Follow along with the demo to learn how to calculate crime rates using R and the Tidyverse. We'll also introduce a new library, janitor, which helps make column names easier to work with. You should get started by loading the tidyverse in a new R block, then install and load janitor.

### Task 1: Load the tidyverse and install and load janitor
```{r}
library(tidyverse)
library(janitor)
```



### Task 2: Load the crime dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv
```{r}
crimes <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv") |> clean_names()
```


### Task 3: Use janitor to clean up the column names



### Task 4: Calculate the total crime rate for each year (picking our unit!)
```{r}
totalRate <- crimes |> mutate(totalRate = (grand_total/population)*10000)
```



### Task 5: Calculate the violent and property crime rates for each year (same unit)
```{r}
specificCrimeRate <- crimes |> mutate(violentRate = (violent_crime_total/population)*10000, propertyRate = (property_crime_total/population))

```



### Task 6: Describe our results

Write a few sentences about what you found in the data. What years had the highest crime rates? What years had the lowest crime rates? Does the pattern match your expectations?

The years with the highest crime rate was 1980. The lowest crime rate was in 2020. The pattern shows that there was a spike in the 90s but crime has been falling since 2001. It matches what I've heard about how we report crime. It's more sensationalized than ever because of the internet but statistically there is just less crime per capita going on. I was surprised that even the raw number of crime was significantly higher in the 90s.


When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.

