---
title: "Prince George's County Car Theft Spike Analysis"
author: "Razak Diallo"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

## Introduction

This analysis examines the impact of a significant spike in car theft rates in 2023 in Prince George's County, and whether that spike has led to a sustained higher rate of car theft in the county. We'll use Analysis of Variance (ANOVA) to determine if theft rates differ significantly between pre-spike (before 2023), spike (2023), and post-spike (2024-present) periods.

## Data Preparation and Exploration

### Loading Libraries and Data

```{r load-libraries}
# Load required libraries
library(tidyverse)
library(lubridate)
```

```{r load-data}
# Import data
theft_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month17_25.csv")

# Look at the first few rows
head(theft_data)

# Get a summary of the data
summary(theft_data)
```


### Get the data how we want it and calculate the rates

**Task 1** Replace REPLACE_ME with the variables you need to calculate the theft_rate (2 points)

```{r process-data}
# Calculate theft rate and create time period variable
theft_data <- theft_data |>
  mutate(
    date = as.Date(month),
    year = year(date),
    month_number = month(date),
    theft_rate = (total / population) * 100000,
    time_period = case_when(
      year < 2023 ~ "Pre-spike",
      year == 2023 ~ "Spike year",
      year > 2023 ~ "Post-spike"
    )
  )

# Create summary by time period
period_summary <- theft_data |>
  group_by(time_period) |>
  summarize(
    mean_thefts = mean(total),
    mean_rate = mean(theft_rate),
    median_rate = median(theft_rate),
    sd_rate = sd(theft_rate),
    n = n(),
    .groups = "drop"
  )

# Display the summary
knitr::kable(period_summary, digits = 2,
             caption = "Summary of Car Theft Rates by Time Period")
```

**Task 2** Describe the results of calculating the rates for each period. What do they suggest? (3 points)

The results suggest that the average theft rate increased during the spike and never returned to prespike levels. But the standard deviation increased overtime as well, indicating that there's a possibility that the increased variability in later years is shifting the average higher. Also the n value is significantly higher for pre-spike levels which could be a bit misleading. If theft was generally trending upwards, the pre-spike average wouldn't properly account for a gradual rise.

## Visualizing the Spike Pattern

### Time Series Plot

```{r time-series}
# Time series plot showing the spike pattern
ggplot(theft_data, aes(x = date, y = theft_rate)) +
  geom_line() +
  geom_smooth(method = "loess", span = 0.2) +
  geom_vline(xintercept = as.Date("2023-01-01"), linetype = "dashed", color = "red") +
  geom_vline(xintercept = as.Date("2024-01-01"), linetype = "dashed", color = "red") +
  annotate("text", x = as.Date("2023-07-01"), y = max(theft_data$theft_rate) * 0.9, 
           label = "Spike Year", color = "red") +
  labs(title = "Monthly Car Theft Rates (2017-2025) with Spike Highlighted",
       x = "Date", y = "Thefts per 100,000 Population") +
  theme_minimal()
```

**Task 3** Describe the output of this chart and then give it a better, more active title that provides the lede (3 points)

The chart shows that there was a sharp increase in theft heading into the spike year after years of consistency. After the spike year, the rate seems to be trending down but still remains above pre-spike levels. 

Title: Monthly car theft rates before, during and after spike


Now let's take a look at average theft rates.

### Yearly Average Theft Rates

```{r yearly-means}
# Calculate yearly means
yearly_means <- theft_data |>
  group_by(year) |>
  summarize(
    mean_rate = mean(theft_rate),
    .groups = "drop"
  )

# Plot yearly means
ggplot(yearly_means, aes(x = year, y = mean_rate)) +
  geom_line(size = 1) +
  geom_point(size = 3, color = "blue") +
  geom_point(data = yearly_means |> filter(year == 2023), 
             aes(x = year, y = mean_rate), size = 4, color = "red") +
  labs(title = "Average Annual Car Theft Rates",
       x = "Year", y = "Average Thefts per 100,000 Population") +
  theme_minimal() +
  scale_x_continuous(breaks = 2017:2025)
```

## One-way ANOVA Analysis

### Hypothesis Formulation

For our ANOVA test, we'll compare the mean theft rates between the three time periods. First we need to create our hypotheses.

**Task 4** Fill in the hypotheses (2 points)

- **Null Hypothesis (H0)**: There is not a significant difference between the car theft rates before, during and after the spike

- **Alternative Hypothesis (H1)**: There is a significant difference between the car theft rates before, during and after the spike

### One-way ANOVA Test

```{r anova-test}
# Perform one-way ANOVA on theft rates by time period
period_anova <- aov(theft_rate ~ time_period, data = theft_data)
summary(period_anova)
```

**Task 5** Describe the result: are there real differences in the average rates between the time periods? Is this result significant? (5 points)

Due to the high F value (218) and low p-value we can reject the null hypothesis and there is a statistically significant difference between the time periods.

## Compare the differences between specific periods

Run our post-anova Tukey test and examine the results. 

```{r posthoc}
# Tukey's HSD test
tukey_results <- TukeyHSD(period_anova)
print(tukey_results)
```
Pre v Post: 20 less thefts
Spike v Post: 17 more thefts
Spike v Pre: 37 more thefts

```{r period-means}
# Visualize time period means with confidence intervals
period_rates <- theft_data |>
  group_by(time_period) |>
  summarize(
    mean_rate = mean(theft_rate),
    se = sd(theft_rate) / sqrt(n()),
    .groups = "drop"
  ) |>
  mutate(
    lower_ci = mean_rate - qt(0.975, n() - 1) * se,
    upper_ci = mean_rate + qt(0.975, n() - 1) * se
  )

# Ensure time period is ordered correctly
period_rates$time_period <- factor(period_rates$time_period, 
                                   levels = c("Pre-spike", "Spike year", "Post-spike"))

ggplot(period_rates, aes(x = time_period, y = mean_rate, fill = time_period)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.2) +
  labs(title = "Car Theft Rates by Time Period with 95% Confidence Intervals",
       x = "", y = "Mean Theft Rate per 100,000") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Conclusion

This analysis has examined whether there are significant differences in car theft rates between the pre-spike period (before 2023), the spike year (2023), and the post-spike period (after 2023).

**Task 6** Describe the results of the Tukey's test and chart showing confidence intervals. Is there a significant difference between the three periods? In particular, is the difference between the pre-spike and post-spike periods real and significant? Write a headline and lede reporting these findings, focusing on the pre- and post-spike periods. (5 points)

With a p-value of 0 for each comparison the Turkey test foubd:
Pre v Post: One average there are 20 more thefts per 100,000 people after the spike when compared to before the spike
Spike v Post: The spike averaged 17 more thefts per 100,000 people when compared to after the spike
Spike v Pre: The spike averaged 37 more thefts per 100,000 people when compared to before the spike

A new normal: Car theft rate has yet to recover from spike

Prince George's County's car theft rate hasn't recovered from the spike in thefts that occurred in 2023. The current average theft rate is 20 incidents above the average before the spike.
