---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
library(dplyr) #I don't know if I did something wrong but the summarize function wasn't working until I did this
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
whSalaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

The first thing I gravitated to was the highest paid positions. Most of them are advisors and it's kind of funny how the second highest paid is a special advisor for economic development. I think it would have been useful to have a separate column for the department the worker belongs to. Then we could get insight into whether staffers dealing with the economy are paid more than those dealing with education and so on. It could be because there is some overlap but it still would've been interesting to look at.

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
whSalaries |> summarize(medianSalary = median(salary))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
whSalaries |> summarize(medianSalary = median(salary), meanSalary = mean(salary), minSalary = min(salary), maxSalary = max(salary))
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean being less than the median tells me that the data is skewed negatively. The minimum and maximum point me to the potential outliers in the data. Considering that the skew is negative, the $0 salaries appear to be the major outliers in the data. 251258 is also an outlier, distancing itself from the second highest salary by ~60 thousand. But the 0 salaries have a higher frequency.

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

This may not end up as anything and could just be a normal thing that the government does but what was the choice to either not pay or, what I think occured here, not report the actual salary of the $0 positions. Moreover, three of these positions deal with women/gender policy. What is going on there? After researching each person by name, I would probably reach out to the employees and see if there is something that is missing from the data that explains them not getting paid. Also I looked up Michelle Barrans (the 250k salary) and found this website called Inside Biden's Basement which is reflective a conceptually good effort to hold people in our government accountable but then the home page is gnarly rants about the swamp. 
