---
title: "HW7_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse
```{r}
library(tidyverse)
```


### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
mm <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```



### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10.
```{r}
sample5 <- mm |> sample_n(5)
sample10 <- mm |> sample_n(10)
```



### Task 4: Calculate the mean and standard deviation of the samples, using one of the colors.
```{r}
sample5 |> 
  summarize(
    mean = mean(red),
    sd = sd(red)
  )
sample10 |> 
  summarize(
    mean = mean(red),
    sd = sd(red)
  )
```



### Task 5: Calculate the mean and standard deviation of the population, using the same color as Task 4
```{r}
mm |> 
  summarize(
    mean = mean(red),
    sd = sd(red)
  )
```


### Task 6: Compare the sample and population statistics

Compare the three sets of means and standard deviations. Are they similar?
The smallest sample is pretty dissimilar from the other two means and standard deviations at 60.2 and 10.5 respectively. The 10 sample has a mean of 69.6 and 19.1 whiel the entire population has a mean of 73 and a standard deviation of 16.5. 

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?
The data tells me that the samples of the samples still trend towards being 50/50. The sample means and sds show that the more data you have the closer you get to the true mean and standard deviation. 


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
