---
title: "final_project"
output: html_document
---

#Load the tidyverse
```{r}
library(tidyverse)
```

#load admissions data
```{r}
hsAdmissions <- read_csv("admissions.csv")
```
#cleaning data (N/A values and <[number])
```{r}
#simplify year
hsAdmissions <- hsAdmissions |> mutate(year = sub("-\\d{2}$", "", year))

#seperate rows with NA values and remove them from the original database
naRows <- hsAdmissions[is.na(hsAdmissions$applied) | is.na(hsAdmissions$admitted) | is.na(hsAdmissions$acceptance_rate), ]
hsAdmissions <- hsAdmissions[!is.na(hsAdmissions$applied) | !is.na(hsAdmissions$admitted) | !is.na(hsAdmissions$acceptance_rate), ]

#rename "Maryland Unknown" entries
hsAdmissions <- hsAdmissions |> mutate(county = ifelse(is.na(county), "Maryland Unknown", county))

# Function to convert "<number" to "number - 0.01"
convert_less_than <- function(value) {
  if (grepl("^<\\d+$", value)) {  # Check if the string starts with "<" followed by digits
    number <- as.numeric(sub("^<", "", value))  # Remove "<" and convert to numeric
    return(number - 0.01)  # Return the number - 0.01
  } else {
    return(as.numeric(value))  # If not, return the value as numeric
  }
}

# Remove "<[number]" values from dataset by converting them to the
hsAdmissions$applied <- sapply(hsAdmissions$applied, convert_less_than)
hsAdmissions$admitted <- sapply(hsAdmissions$admitted, convert_less_than)
hsAdmissions$acceptance_rate <- sapply(hsAdmissions$acceptance_rate, convert_less_than)
```
If I had kept the N/A values I wouldn't have been able to operate on the data set. I thought about turning them to 0s but there could be numerous reasons as to why the data is unavailable such as it simply not being tracked. I also changed the <[number] values to the highest they could possibly be without crossing the threshold. This gives the most charitable data to the university's argument that since 2020 it has "broaden our student recruitment efforts across Maryland counties and cities." My conclusion will consider this decision.

#Determining the number of N/A years per county
```{r}
naByCounty <- naRows %>%
  group_by(county) %>%
  summarise(na_count = n())

```
It's interesting how the NA values are overwhelmingly in Prince George's County. If the N/A were to represent schools where 0 applicants applied to the University of Maryland, it shows a strong lack of interest in the institution.

#Admissions rate of each county per year. Percentage of total applications and total admitted by county per year
```{r}
# Calculate the percentage of admitted students per county for each year
countyAdmissionsRate <- hsAdmissions %>%
  group_by(year, county) %>%
  summarise(
    applied = ceiling(sum(applied, na.rm = TRUE)),
    admitted = ceiling(sum(admitted, na.rm = TRUE)),
    admissionsRate = (admitted/applied)*100,
    .groups = 'drop'
    ) |> 
  group_by(year) |> 
  mutate(totalAppliedYear = sum(applied), totalAdmittedYear = sum(admitted)) %>%
  mutate(perctApplied = (applied / totalAppliedYear)*100, perctAdmitted = (admitted / totalAdmittedYear) * 100)
```

There are a few interesting things here. On the admissions rate front you would expect I applicant counties to fall closer to the mean at just under 60% but a high volume of years for Anne Arundel County and Prince George's County (both which saw application rates in the thousands) are as low as 39%.
```{r}
#summary statistics
arSummary <- countyAdmissionsRate |> 
  group_by(county) |>
  summarize(
    meanAdmissions = mean(admissionsRate),
    medianAdmissions = median(admissionsRate),
    sdAdmissions = sd(admissionsRate),
    minAdmissions = min(admissionsRate),
    maxAdmissions = max(admissionsRate)
  )

print(arSummary)

countyAdmissionsRate |> 
  ggplot() + 
  geom_histogram(aes(x = admissionsRate), binwidth = 6) +
  geom_vline(aes(xintercept = mean(admissionsRate)), color = "red", linetype = "dashed", size = 1) + 
  geom_vline(aes(xintercept = mean(admissionsRate) - sd(admissionsRate)), color="blue", linetype = "dashed", size =1) +
  geom_vline(aes(xintercept = mean(admissionsRate) + sd(admissionsRate)), color="blue", linetype = "dashed", size =1)
```
The shifts in the proportion of applicants to admitted also raises a few questions. The majority of county/years see increases or negligible changes in their share after admission, except for Anne Arundel County and Prince George's County who average large decreases in their share of the pool. Montgomery County exist on the opposite, with an average increase close to 2 percentage points after admission.
```{r}
changesInPercentage <- countyAdmissionsRate |> 
  group_by(county) |> 
  summarize(
    change = mean(perctAdmitted - perctApplied)
  )
```


#narrowing in on before and after 2020
```{r}
prePost2020 <- countyAdmissionsRate |> 
  group_by(county) |> 
  summarize(
    meanAdmissionsRatePre = mean(admissionsRate[year<=2020]),
    meanAdmissionsRatePost = mean(admissionsRate[year>2020]),
    meanApplicationPerctPre = mean(perctApplied[year<=2020]),
    meanApplicationPerctPost = mean(perctApplied[year>2020]),
    meanAdmittedPerctPre = mean(perctAdmitted[year<=2020]),
    meanAdmittedPerctPost = mean(perctAdmitted[year>2020]),
    meanChangePre = mean(perctAdmitted[year<=2020] - perctApplied[year<=2020]),
    meanChangePost = mean(perctAdmitted[year>2020] - perctApplied[year>2020])
  )
```
The data here shows increases in admissions rate across the board. Anne Arundel county shows a decrease in their share of admitted students after the 2020 while Prince George's County managed to increase. 

#T-Test
Testing if claim that across the board admissions broadened is true by looking at the percentage make up of the admitted class.

- H₀: μ ≤ 4.519774 (Recruiting efforts have not broadened population)
- H₁: μ > 4.519774 (Recruiting efforts have broadened population)
```{r}
before_2020 <- countyAdmissionsRate %>% filter(year <= 2020)
after_2020 <- countyAdmissionsRate %>% filter(year > 2020)

t_test_result <- t.test(after_2020$perctAdmitted, mu=mean(before_2020$perctAdmitted), alternative = "greater")

print("T-Test Results:")
print(t_test_result)

# Extract the p-value
p_value <- t_test_result$p.value

# Calculate the 95% confidence interval
ci <- t.test(after_2020$perctAdmitted)$conf.int

# Format and display the confidence interval in a readable way
ci_formatted <- tibble(
  `Lower bound` = ci[1],
  `Upper bound` = ci[2],
  `Confidence level` = "95%"
)

# Display as a nice table
ci_formatted

# Set our significance level
alpha <- 0.05

# Compare and make a decision
if(p_value < alpha) {
  conclusion <- "The data provides statistical evidence supporting the claim that the recruitment efforts have broadned the in-state population."
} else {
  conclusion <- "The data does not provide strong statistical evidence that the recruitment efforts have broadned the in-state population."
}

# Create a tibble to display the key statistics for your story
story_stats <- tibble(
  `Previous average` = mean(before_2020$perctAdmitted),
  `New average` = mean(after_2020$perctAdmitted),
  `Decrease` =  mean(after_2020$perctAdmitted) - mean(before_2020$perctAdmitted),
  `Percent change` = paste0(round(((mean(after_2020$perctAdmitted) - mean(before_2020$perctAdmitted)) / mean(before_2020$perctAdmitted)) * 100, 1), "%"),
  `p-value` = p_value,
  `Statistically significant?` = if_else(p_value < alpha, "Yes", "No")
)

# Display the key statistics
story_stats

print(conclusion)
```

The t-test of the entire population, without filtering by county, shows that recruitment efforts did not dramatically shift the overall distribution of in-state students by county. You could argue that a homogeneous group of students was not the goal of the University but rather increasing recruitment in certain counties. 

#Individual County T-Test
```{r}
# Create a results dataframe
results <- data.frame(county = character(),
                      t_statistic = numeric(),
                      p_value = numeric(),
                      mean_before = numeric(),
                      mean_after = numeric(),
                      stringsAsFactors = FALSE)

# Get unique counties
unique_counties <- unique(hsAdmissions$county)

# Loop through each county and perform t-test
for (c in unique_counties) {
  before_data <- before_2020 %>% filter(c == county)
  after_data <- after_2020 %>% filter(c == county)
  
  # Check if there are enough data points to perform the t-test
  if (nrow(before_data) > 1 && nrow(after_data) > 1) {
    t_itest_result <- t.test(after_data$perctAdmitted, mu=mean(before_data$perctAdmitted), alternative = "greater")
    
    # Store the results
    results <- rbind(results, data.frame(
      county = c,
      t_statistic = t_itest_result$statistic,
      p_value = t_itest_result$p.value,
      mean_before = mean(before_data$perctAdmitted, na.rm = TRUE),
      mean_after = mean(after_data$perctAdmitted, na.rm = TRUE)
    ))
  }
}

# View the results
print("T-Test Results by County:")
print(results)
```

Most of the counties had statistically insignificant results. Prince George's County is close to the 0.05 threshold and has a clear jump in its mean share of admitted students by 1 percentage point. Nonetheless because none of the counties meet the standard alpha value, the results are statistically significant enough to conclude that the recruitment efforts by the University have been effective.

#Story Pitch
The results show statistically insignificant change in share of total admitted students. Also considering how charitable the data was to the University (eliminating N/A and rounding up <[number]), the story could discuss what UMD's broadened recruitment approach has been and what other factors could be contributing to to it's insignificant change in the composition of the university. 

A major problem with my analysis is my approach to cleaning up the data. Someone else could do it entirely differently, considering N/A and <[number] as 0 for example, and get super different results. Moreover there could be better determinants of recruitment than changes in percentage of admitted student population. Some counties are just bigger/smaller than others and it would be unrealistic for the demographics to change that dramatically. Also the data ends only two years after 2020 and some recruitment efforts may not fully trickle down until 2024 or 2025. 

If I was to continue reporting on this story I would aim to get updated data from the university that groups all admissions by county (to avoid anonymizing for smaller populations). I would spend more time with the data, looking at z-scores to see if there were any outlier years for specific counties. I would also speak to the University and see what efforts they specifically made to broaden their recruitment. I think it would also be interesting to consider where a lot of transfer students from Community Colleges that graduated high school in the past 4 years are coming from.








